He-Man: A domain-specific language for lightweight concurrency

### 2014-04-02 paper/paper.tex

\lstset{
  language=Haskell,
  basicstyle=\small\ttfamily,
  frame=trbl,
  basewidth={0.5em,0.45em},
  commentstyle=\ttfamily\color{ForestGreen},
  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
            {\\\\}{{\char`\\\char`\\}}1
            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
            {>>}{{>>}}2 {>>=}{{>>=}}2
            {|}{{$\mid$}}1
}
\lstnewenvironment{code}{\lstset{frame=}}{}

@@

@@

%XXX: new title?
\title{He-Man: A domain-specific language for lightweight concurrency}
%\subtitle{I said hey, what's goin' on}
\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}

@@

\begin{abstract}
The event-driven model is the gold standard for highly concurrent network
application programming; while threaded code is easier to write and reason
about, it is considerably less scalable. We observe that these paradigms are not
only interchangeable, but primarily confined to an application's core control
flow; we propose a hybrid model in which programmers express this core logic at
a high level of abstraction, yet are free to implement the remainder of the
application at a low level.

@@

Modern network servers, like HTTP, IRC, and multiplayer games, need to support
many simultaneous, possibly idle, connections---this is known as the C10K
(``10,000 client'') problem \cite{Kegel}. Because I/O latency is high,
particularly over a network, our goal is to process ready clients while others
are blocked on I/O. To achieve maximum throughput, such servers are usually
written in low-level languages like C.

@@

However, as we discuss in the next section, these code transformations are
familiar to functional programmers as closure conversion (to access connection
state across function boundaries), continuation-passing style (via function
pointers) with trampolining (yielding to the event loop after registering I/O).
These are well-studied conversions which can be performed mechanically by the
He-Man backend.

@@

As an example of the power that Haskell and the monadic style gives
us, Figure \ref{fig:generic} shows our library routines for simulating
blocking network operations. \t{do\_nb\_action} takes an
\t{Expr} and a \t{Prog IntE} action representing a network call that
can return \t{EAGAIN} and produces an action that generates code to
simulate a blocking network operation. This allows us to describe how
to simulate blocking I/O in a generic way, and then invoke it to
create our concrete wrappers. (\t{.=.} is a wrapper for \b{Assign} 
that takes a \t{Prog Expr} as its second argument and the \t{sock\_*} 
functions invoke \b{Call} wrappers on the appropriate network calls.)

@@

TODO here we will discuss the transformations that occur

%As an illustrative example, consider encountering an \b{If} statement in our
%backwards pass, where we have access to the translation of the later
%statements, as well as the earlier statements. Morally, we would like to
%translate this to a conditional jump to the translation of either the first or
%second arm, each of which is terminated by a jump to the later statements.
%(Note that \t{flattenPrgm} leaves all optimization to the next pass.)

%We package up the later statements in a new block, and then translate the
%preceding statements, terminated by a tail-position conditional \b{If}. We also
%translate each arm of the original \b{If}, terminated by a tail-position
%\b{Goto} to the block containing the later statements. Importantly, all three
%of these recursive translations may generate new blocks, which will be
%collected automatically by the \t{Writer} monad.

%Because blocking I/O may only happen in the main loop, we also terminate a
%block after any blocking I/O call, by inserting a \b{GotoWait} to the block
%containing the later statements. \b{GotoWait} is identical to \b{Goto} except
%to signal to later passes that the thread \emph{must} yield to the main loop at
%this location, and thus the goto must not be optimized away.
% TODO for concurrency-related reasons

@@

At this point, we need to eliminate spurious control flow edges from our state
machine in order to enable further block elimination. To do this, we perform
block-local constant propagation and folding, and dead code elimination. At face
value, it is not important for us to optimize straight-line code: our output is
further optimized by a C compiler, and execution time is dominated by I/O, not
computation. However, simple constant folding eliminates many \t{if} statements
%TODO examples, justification

TODO here we discuss the fuseBlocks pass

@@

The He-Man version is 33 lines long and clearly represents the control flow of
each individual thread. (This does not include the He-Man library, which has 25
lines of code implementing generic routines for simulating blocking I/O using
nonblocking I/O and wait.) The total line counts, somewhat understate the
difference between the versions, since the boilerplate for setting up
connections is about the same for both versions. When just looking at the core
read/write loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

@@

% XXX work on the below paragraphs

@@

It is important to note that in environments providing cooperative
multithreading, like Capriccio, FoxNet, and He-Man, while the runtime guarantees
that no thread monopolizes execution while waiting for I/O, it is certainly
possible for one thread to monopolize execution by running indefinitely.
Bounding execution time is a difficult problem which has been studied in the
context of grid computing \cite{VanderwaartCrary}. The He-Man compiler
guarantees that no code block contains an infinite loop; unbounded execution can
only occur in calls to external functions.

>>>>

@@

\lstset{
  language=Haskell,
  basicstyle=\small\ttfamily,
  frame=trbl,
  basewidth={0.5em,0.45em},
  commentstyle=\ttfamily\color{ForestGreen},
  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
            {\\\\}{{\char`\\\char`\\}}1
            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
            {>>}{{>>}}2 {>>=}{{>>=}}2
            {|}{{$\mid$}}1
}
%\lstnewenvironment{code}{\lstset{frame=}}{}
\newenvironment{code}{\endgraf\small\verbatim}{\endverbatim}

@@

\title{He-Man: A domain-specific language for lightweight concurrency}
%\subtitle{I said hey, what's goin' on}
\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}

@@

\begin{abstract}
The event-driven model is the gold standard for highly concurrent network
application programming; while threaded code is easier to write and reason
about, it is considerably less scalable. We observe that these paradigms are not
only interchangeable, but primarily confined to an application's core I/O
control flow logic. We propose a hybrid model in which programmers express this
core logic in a high-level concurrency abstraction, yet are free to implement
the remainder of the application in a generic fashion.

@@

Modern network servers, as for HTTP, IRC, and multiplayer games, need to support
many simultaneous, possibly idle, connections---this is known as the C10K
(``10,000 client'') problem \cite{Kegel}. Because I/O latency is high,
particularly over a network, our goal is to process ready clients while others
are blocked on I/O. To achieve maximum throughput, such servers are usually
written in low-level languages like C.

@@

However, as we discuss in the next section, these code transformations are
familiar to functional programmers as closure conversion (to access connection
state across function boundaries), continuation-passing style (via function
pointers) with trampolining (yielding to the event loop after registering I/O).
These are well-studied conversions which can be performed mechanically by the
He-Man back-end.

@@

As an example of the power that Haskell and the monadic style gives
us, Figure \ref{fig:generic} shows our library routines for simulating
blocking network operations. \t{do\_nb\_action} takes an
\t{Expr} and a \t{Prog IntE} action representing a network call that
can return \t{EAGAIN} and produces an action that generates code to
simulate a blocking network operation. This allows us to describe how
to simulate blocking I/O in a generic way, and then invoke it to
create our concrete wrappers. (\t{.=.} is a wrapper for \b{Assign} 
that takes a \t{Prog Expr} as its second argument, and the \t{sock\_*} 
functions invoke \b{Call} wrappers on the appropriate network calls.)

@@

\b{If} statements are the most interesting case. We place the subsequent
statements in a separate block, and process the two bodies of the \b{If}
statement, placing each in a separate block, and terminating each with a tail
\b{Goto} to the subsequent statements. We then process the statements preceding
the \b{If}, terminating their block with a tail-position \b{If} which jumps to
either branch. \b{While} statements are handled in a similar fashion.

We translate thread-blocking operations ($\b{Wait}~e$ in the front-end language)
to a blocking jump ($\b{GotoWait}~e$) to a new code block containing the next
statements to execute. \b{GotoWait} is similar to \b{Goto}, except that it
contains an event to wait on, and is guaranteed not to be optimized away by
later passes. Thread spawns are translated into a \b{Spawn} statement which
takes a list of arguments and a destination label; the thread's code, which is
part of the front-end \b{Spawn} statement, is processed separately.

@@

At this point, we would like to eliminate spurious control flow from our state
machine in order to enable further block elimination. Our \t{optimize} pass
performs block-local constant propagation and folding, symbolic constraint
propagation, and dead code elimination. At face value, it is not important for
us to optimize straight-line code: our output is further optimized by a C
compiler, and execution time is dominated by I/O, not computation. However,
these simple optimizations can statically resolve many \b{If} tails generated by
the previous pass, resulting in substantially less control flow.

Consider, for example, the following code snippet from our library function
\t{full\_write}:
\begin{smallverbatim}
full_write (fd, e) buf size = do
  amt_written <- var "total_written" Int 0
  failed <- var "failed" Int 0
  while (amt_written .< size .&& failed .== 0) $ do
    -- write amt bytes to buffer
    ifE (amt .<= 0) (failed .= 1) $ -- amt_written += amt
\end{smallverbatim}
After only \t{simplifyJumps}, this function compiles to eight blocks, including
the following:
\begin{smallverbatim}
Block 9 {
  IInt total_written_16 = 0
  IInt failed_17 = 0
  if (total_written_16 < file_size_10 && failed_17 == 0)
    goto 8
  else
    goto 7
}
Block 40 {
  failed_17 := 1
  if (total_written_16 < file_size_10 && failed_17 == 0)
    goto 8
  else
    goto 7
}
\end{smallverbatim}
Block 9 is executed the first time the \t{while} loop is encountered; block 40
is executed on the \t{amt .<= 0} failure condition, and jumps back to the top
of the \t{while} loop. But constant folding shows that the loop is always
executed at least once, so long as the file size is positive, and is always
exited when the failure condition is tripped. After optimization, we are left
with:
\begin{smallverbatim}
Block 9 {
  IInt total_written_16 = 0
  IInt failed_17 = 0
  if (0 < file_size_10)
    goto 8
  else
    goto 7
}
Block 40 {
  failed_17 := 1
  goto 7
}
\end{smallverbatim}
While \t{optimize} does not result in much smaller code, in practice it does
remove many edges from the control flow graph.

For performance reasons, we would like to yield to the event loop sparingly,
which amounts to increasing the length of each block as much as possible. Our
final optimization pass, \t{fuseBlocks}, hoists certain blocks into their
callers to eliminate unnecessary yields. To facilitate hoisting, we generalize
the grammar to allow statements in each branch of an \b{If} tail.

We hoist blocks which are invoked in a unique location, and those which end in a
thread \b{Exit} or blocking I/O (\b{GotoWait}). The former eliminates yields
without increasing code size; the latter allows threads to register I/O or free
resources immediately. In the \t{full\_write} example, block 40 is only reachable
from the failure \t{if}, and is eliminated entirely, while the thread exit
located after the \t{while} loop is hoisted into block 9.
\begin{smallverbatim}
Block 9 {
  IInt total_written_16 = 0
  IInt failed_17 = 0
  if (0 < file_size_10)
    goto 8
  else
    IInt tmp_19 = close(tmp_9)
    exit
}
\end{smallverbatim}
After optimization, \t{full\_write} is reduced from eight to four blocks.

@@

The He-Man version is 33 lines long and clearly represents the control flow of
each individual thread. (This does not include the He-Man library, which has 25
lines of code implementing generic routines for simulating blocking I/O using
nonblocking I/O and wait.) The total line counts understate the
difference between the versions, since the boilerplate for setting up
connections is about the same for both versions. When just looking at the core
read/write loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

@@

% XXX work on the below paragraphs?

@@

Cooperative multithreading (as in Capriccio, FoxNet, and He-Man) is more
performant in this domain than preemption, because it avoids unnecessary context
switches and executes only unblocked threads. On the other hand, the scheduler
is only invoked when a thread voluntarily yields, so a thread may monopolize
execution time if it never yields. Placing bounds on execution time is a
difficult problem in general, and has been studied in the context of grid
computing \cite{VanderwaartCrary}. The He-Man compiler guarantees that no code
block contains an infinite loop; unbounded execution can only occur in calls to
external functions.
